---
title: "Predictive Modeling Report"
author: "Giodanno Limin"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(mgcv)
```

```{r 0}
# get the training data
train <- read.csv("trainingdata.csv")

# direct access
attach(train)
# write a cv_rmse function
cv_rmse <- function(formula, data, K = 10) {
  set.seed(1)
  n <- nrow(data)
  folds <- sample(rep(1:K, length.out = n))
  
  rmse <- numeric(K)
  
  for (k in 1:K) {
    train_fold <- data[folds != k, ]
    test_fold  <- data[folds == k, ]
    
    fit <- lm(formula, data=train_fold)
    pred <- predict(fit, newdata=test_fold)
    
    rmse[k] <- sqrt(mean((pred - test_fold$y)^2))
  }
  
  return(list(
    mean_RMSE = mean(rmse),
    sd_RMSE   = sd(rmse),
    all_RMSE  = rmse
  ))
}

```

```{r 1}
lm1 = lm(y~.,data=train)
summary(lm1)
cv_rmse(lm1,train)
```
```{r 2}
sig_vars <- summary(lm1)$coefficients
sig_vars[sig_vars[, "Pr(>|t|)"] < 0.05, ]
lm2 = lm(y~X7+X9+X19+X21+X30+X40+X87+X88+X111,data=train)
summary(lm2)
cv_rmse(lm2,train)
```
Let's try better way
```{r 4}
cor_with_y <- function(x) {
  cor(train$y, x)
}
cors <- sapply(train[, -1], cor_with_y)
head(sort(abs(cors), decreasing=TRUE), 10)

top10 <- names(sort(abs(cors), decreasing = TRUE))[1:10]
top10

for (var in top10) {
  plot(
    train[[var]], train$y,
    main = paste(var, "vs y"),
    xlab = var,
    ylab = "y",
    pch = 19,
    col = rgb(0, 0, 1, 0.4)   
  )
}

lm4 = lm(y~X66+X69+X21,data=train)
summary(lm4)
cv_rmse(lm4,train)
```
Now let's check if the adding other variable lower the mean_RMSE
"X19"  "X45"  "X111" "X41"  "X38"  "X3"   "X59"
```{r 5}
vars_to_test <- c("X19", "X45", "X111", "X41", "X38", "X3", "X59")

results <- data.frame(
  variable = vars_to_test,
  mean_RMSE = NA
)

base_form <- "y ~ X66 + X69 + X21"
base_rmse <- cv_rmse(as.formula(base_form), train)$mean_RMSE

results <- data.frame(
  variable = vars_to_test,
  mean_RMSE = NA,
  improvement = NA
)

for (i in seq_along(vars_to_test)) {

  var <- vars_to_test[i]

  # construct formula dynamically
  form <- as.formula(
    paste(base_form, "+", var)
  )

  # compute rmse
  new_rmse <- cv_rmse(form, train)$mean_RMSE

  # fill results
  results$mean_RMSE[i] <- new_rmse
  results$improvement[i] <- new_rmse < base_rmse
}

results

lm5 = lm(y~X66+X69+X21+X19+X111+X3+X59,data=train)
cv_rmse(lm5,train)
```
Now let's try polynomial for X66, X69, X21, and X3.
```{r 6}
search_poly_multivar <- function(max66 = 10, max69 = 10, max21 = 10) {
  results <- data.frame()
  
  for (d66 in 1:max66) {
    for (d69 in 1:max69) {
      for (d21 in 1:max21) {
        
        form <- as.formula(
          paste(
            "y ~ poly(X66,", d66, ",raw=TRUE) +",
            "poly(X69,", d69, ",raw=TRUE) +",
            "poly(X21,", d21, ",raw=TRUE) +",
            "+ X19 + X111 + X3 + X59",
            sep=" "
          )
        )
        
        cv <- cv_rmse(form, train)$mean_RMSE
        
        results <- rbind(
          results,
          data.frame(
            d66 = d66,
            d69 = d69,
            d21 = d21,
            RMSE = cv
          )
        )
      }
    }
  }
  
  return(results)
}

# run the search
poly_results <- search_poly_multivar()
best_row <- poly_results[which.min(poly_results$RMSE), ]
best_row

search_poly_X3 <- function(max3=10) {
  results <- data.frame()
  
  for (d3 in 1:max3){
        
    form <- as.formula(
      paste(
        "y ~",
        "poly(X66, 5, raw=TRUE) +",
        "poly(X69, 3, raw=TRUE) +",
        "poly(X21, 10, raw=TRUE) +",
        "poly(X3,", d3, ", raw=TRUE) +",
        "X19 + X111 + X59",
        sep=" "
      )
    )
    
    cv <- cv_rmse(form, train)$mean_RMSE
    
    results <- rbind(
      results,
      data.frame(
        d3 = d3,
        RMSE = cv
      )
    )
  }
  
  return(results)
}

# run the search
poly_results <- search_poly_X3()

best_row <- poly_results[ which.min(poly_results$RMSE) , ]
best_row

# let's check again for X21, since previously our max degree is 10,
# maybe if the degree greater than 10 is better

search_poly_X21 <- function(max21 = 20) {
  results <- data.frame()
  
  for (d21 in 10:max21) {
    
    form <- as.formula(
      paste(
        "y ~",
        "poly(X66, 5, raw=TRUE) +",
        "poly(X69, 3, raw=TRUE) +",
        "poly(X3,  3, raw=TRUE) +",
        "poly(X21,", d21, ", raw=TRUE) +",
        "X19 + X111 + X59",
        sep=" "
      )
    )
    
    cv <- cv_rmse(form, train)$mean_RMSE
    
    results <- rbind(results, data.frame(d21=d21, RMSE=cv))
  }
  
  return(results)
}
poly21_results <- search_poly_X21(max21 = 15)

# best degree
best_row <- poly21_results[which.min(poly21_results$RMSE), ]
best_row

lm6 = lm(y~poly(X66, 5, raw=TRUE)+
           poly(X69, 3, raw=TRUE)+
           poly(X21, 14, raw=TRUE)+
           poly(X3, 3, raw=TRUE)+
           + X19 + X111 + X59)
cv_rmse(lm6,train)

```
Now let's check the interactions for X66, X69, X21, X3
```{r 7}
base_rmse <- cv_rmse(lm6, train)$mean_RMSE

base_string <- paste(
  "y ~",
  "poly(X66, 5, raw=TRUE) +",
  "poly(X69, 3, raw=TRUE) +",
  "poly(X21,14, raw=TRUE) +",
  "poly(X3,  3, raw=TRUE) +",
  "+ X19 + X111 + X59"
)

candidates <- c(
  "X66:X69",
  "X66:X21",
  "X66:X3",
  "X69:X21",
  "X69:X3",
  "X21:X3"
)

test_interactions <- function(base_string, candidates, train) {
  
  results <- data.frame()
  
  for (int in candidates) {
    full_form_string <- paste(base_string, "+", int)
    form <- as.formula(full_form_string)
    
    cv <- cv_rmse(form, train)$mean_RMSE
    
    results <- rbind(
      results,
      data.frame(
        interaction = int,
        RMSE = cv
      )
    )
  }
  
  return(results)
}
interaction_results <- test_interactions(base_string, candidates, train)
improved <- interaction_results$RMSE < base_rmse
interaction_results[ improved , ]

lm7 = lm(y~poly(X66, 5, raw=TRUE)+
           poly(X69, 3, raw=TRUE)+
           poly(X21, 14, raw=TRUE)+
           poly(X3, 3, raw=TRUE)+
           +X66:X21+X69:X21+X21:X3+
           +X19+X111+X59)
cv_rmse(lm7,train)
```
Next let's try 2nd degree of interactions ()

```{r 8}
generate_interactions2 <- function(var1, var2) {
  terms <- c()
  
  for (i in 1:2) {
    for (j in 1:2) {
      terms <- c(terms, paste0("I(", var1, "^", i, " * ", var2, "^", j, ")"))
    }
  }
  
  return(terms)
}

int_X66_X21 <- generate_interactions2("X66", "X21")
int_X69_X21 <- generate_interactions2("X69", "X21")
int_X21_X3  <- generate_interactions2("X21", "X3")

interaction_list <- c(int_X66_X21, int_X69_X21, int_X21_X3)
length(interaction_list)   # should be 12

# This is lm6, not lm7 to check whether higher interactions are better.

base_string <- paste(
  "y ~",
  "poly(X66,5,raw=TRUE) +",
  "poly(X69,3,raw=TRUE) +",
  "poly(X21,14,raw=TRUE) +",
  "poly(X3,3,raw=TRUE) +",
  "X19 + X111 + X59 "
)


test_high_interactions <- function(base_string, interaction_list, train) {
  
  results <- data.frame()
  
  for (int in interaction_list) {
    
    form_string <- paste(base_string, "+", int)
    form <- as.formula(form_string)
    
    cv <- cv_rmse(form, train)$mean_RMSE
    
    results <- rbind(
      results,
      data.frame(
        interaction = int,
        RMSE = cv
      )
    )
  }
  
  return(results)
}


results_deg <- test_high_interactions(base_string, interaction_list, train)

# Sort by RMSE
sorted_interactions <- results_deg[order(results_deg$RMSE), ]


rmse_lm6 <- cv_rmse(lm6, train)$mean_RMSE

better <- sorted_interactions[ sorted_interactions$RMSE < rmse_lm6 , ]

cat("\n=== Interactions BETTER than lm6 ===\n")
print(better)

better_terms <- c(
  "I(X66^2 * X21^2)",
  "I(X66^2 * X21^1)",
  "I(X66^1 * X21^2)",
  "I(X66^1 * X21^1)",
  "I(X69^1 * X21^2)",
  "I(X69^1 * X21^1)",
  "I(X21^1 * X3^2)",
  "I(X21^2 * X3^2)",
  "I(X21^1 * X3^1)"
)

# best subset 2^9=512 subsets, it's still doable

best_subset <- function(base_string, terms, train) {
  
  best_rmse <- Inf
  best_model <- NULL
  best_terms <- NULL
  
  k <- length(terms)
  
  for (size in 0:k) {
    
    if (size == 0) {
      subsets <- list(character(0))
    } else {
      subsets <- combn(terms, size, simplify = FALSE)
    }
    
    for (subset in subsets) {
      
      # Build formula
      if (length(subset) == 0) {
        form_string <- base_string
      } else {
        form_string <- paste(base_string, "+", paste(subset, collapse = " + "))
      }
      
      rmse <- cv_rmse(as.formula(form_string), train)$mean_RMSE
      
      if (rmse < best_rmse) {
        best_rmse <- rmse
        best_model <- form_string
        best_terms <- subset
        message("New best RMSE: ", rmse,
                " | terms: ", paste(subset, collapse=", "))
      }
    }
  }
  
  list(
    best_rmse = best_rmse,
    best_terms = best_terms,
    best_model = best_model
  )
}


result_bs <- best_subset(base_string, better_terms, train)

result_bs$best_rmse
result_bs$best_terms
result_bs$best_model


lm8 = lm(y~poly(X66, 5, raw=TRUE)+
           poly(X69, 3, raw=TRUE)+
           poly(X21, 14, raw=TRUE)+
           poly(X3, 3, raw=TRUE)+
           +I(X66^2 * X21^2) + I(X69^1 * X21^2) + I(X21^1 * X3^2)+
           +X19 + X111+X59)
summary(lm8)
cv_rmse(lm8,train)
```
Now our model is quite complex. Let's see if removing any of the term
could reduce cv rmse.
```{r 9}
blocks <- c(
  "poly(X66,5,raw=TRUE)",
  "poly(X69,3,raw=TRUE)",
  "poly(X21,14,raw=TRUE)",
  "poly(X3,3,raw=TRUE)",
  "I(X66^2 * X21^2)",
  "I(X69^1 * X21^2)",
  "I(X21^1 * X3^2)",
  "X19",
  "X111",
  "X59"
)

best_subset_blocks <- function(block_terms, train) {
  
  best_rmse  <- Inf
  best_model <- NULL
  best_terms <- NULL
  
  k <- length(block_terms)
  
  for (size in 1:k) {# or 0:k if you want intercept-only too
    subsets <- combn(block_terms, size, simplify = FALSE)
    
    for (subset in subsets) {
      rhs <- paste(subset, collapse = " + ")
      form_string <- paste("y ~", rhs)
      
      rmse <- cv_rmse(as.formula(form_string), train)$mean_RMSE
      
      if (rmse < best_rmse) {
        best_rmse  <- rmse
        best_model <- form_string
        best_terms <- subset
        message("New best RMSE: ", rmse,
                " | terms: ", paste(subset, collapse = ", "))
      }
    }
  }
  
  list(
    best_rmse  = best_rmse,
    best_terms = best_terms,
    best_model = best_model
  )
}

result_blocks <- best_subset_blocks(blocks, train)
result_blocks$best_rmse
result_blocks$best_terms
result_blocks$best_model
lm9 = lm(y ~ poly(X66,5,raw=TRUE) +
           poly(X69,3,raw=TRUE) +
           poly(X21,14,raw=TRUE) +
           poly(X3,3,raw=TRUE) +
           I(X66^2 * X21^2) +
           I(X69^1 * X21^2) +
           I(X21^1 * X3^2)+ 
           X19 + X111)
summary(lm9)
cv_rmse(lm9,train)

# Now we have finished everything with lm. Let's try it on Kaggle.

test <- read.csv("test_predictors.csv")

pred <- predict(lm9, newdata = test)
sample <- read.csv("SampleSubmission.csv")

submission <- data.frame(
  id = sample$id,
  y  = pred
)

write.csv(submission, "LM9_Submission.csv", row.names=FALSE)

cat("Saved: LM9_Submission.csv\n")
```

Score on Kaggle:
Private Score: 0.52147
Public Score: 0.52681

Now let's use GAM to get lower RMSE

s: smoothing splines
ti: smooth interactions
```{r gam 1}
library(mgcv)

# This is similar with LM9, 
# but we change the one variable polynomials to smoothing splines
gam1_formula <- y ~
  s(X66, k = 12) +
  s(X69, k = 12) +
  s(X21, k = 18) +
  s(X3,  k = 10) +
  I(X66^2 * X21^2) +
  I(X69^1 * X21^2) + 
  I(X21^1 * X3^2)+ 
  X19 + X111

# Fit GAM1
gam1 <- gam(gam1_formula, data=train, method="REML", select=TRUE)

summary(gam1)

# Create cv_gam function
cv_gam <- function(formula, data, K = 10, seed = 1) {
  set.seed(seed)
  
  n <- nrow(data)
  folds <- sample(rep(1:K, length.out = n))
  
  rmse <- numeric(K)
  
  for (k in 1:K) {
    train_fold <- data[folds != k, ]
    test_fold  <- data[folds == k, ]
    
    # Fit GAM
    fit <- gam(formula, data = train_fold, method = "REML")
    
    # Predict
    pred <- predict(fit, newdata = test_fold)
    
    # RMSE
    rmse[k] <- sqrt(mean((pred - test_fold$y)^2))
  }
  
  return(list(
    mean_RMSE = mean(rmse),
    sd_RMSE   = sd(rmse),
    all_RMSE  = rmse
  ))
}
cv_gam1 = cv_gam(gam1_formula,train)
cv_gam1
```
Then, we change the interactions to smooth interactions.
```{r gam 2}
gam2_formula <- y ~
  s(X66, k = 12) +
  s(X69, k = 12) +
  s(X21, k = 18) +
  s(X3,  k = 10) +
  ti(X66,X21, k=8) +
  ti(X69,X21, k=8) + 
  ti(X21,X3, k=8)+ 
  X19 + X111

gam2 <- gam(gam2_formula, data=train, method="REML", select=TRUE)

summary(gam2)

cv_gam2 = cv_gam(gam2_formula,train)
cv_gam2
cv_gam2$mean_RMSE < cv_gam1$mean_RMSE
```
From the summary function, the ti(X69,X21) p-value is quite suspicious.
Let's try removing it and check the RMSE.
```{r gam 3}
gam3_formula <- y ~
  s(X66, k = 12) +
  s(X69, k = 12) +
  s(X21, k = 18) +
  s(X3,  k = 10) +
  ti(X66, X21, k=8) +
  ti(X21, X3,  k=8) +
  X111 + X19

gam3 <- gam(gam3_formula, data=train, method="REML", select=TRUE)

summary(gam3)

cv_gam3 = cv_gam(gam3_formula,train)
cv_gam3
cv_gam3$mean_RMSE < cv_gam2$mean_RMSE
```

X19 is a discrete variable, treating it as a categorical variable may improve the GAM’s performance.

```{r gam 4}
train$X19_f <- factor(train$X19)
test$X19_f  <- factor(test$X19, levels = levels(train$X19_f))

gam4_formula <- y ~
  s(X66, k = 12) +
  s(X69, k = 12) +
  s(X21, k = 18) +
  s(X3,  k = 10) +
  
  ti(X66, X21, k=8) +
  ti(X21, X3,  k=8) +
  
  X111 + X19_f

gam4 <- gam(gam4_formula, data=train, method="REML", select=TRUE)

summary(gam4)

cv_gam4 = cv_gam(gam4_formula,train)
cv_gam4
cv_gam4$mean_RMSE < cv_gam3$mean_RMSE

pred <- predict(gam4, newdata = test)
sample <- read.csv("SampleSubmission.csv")

submission <- data.frame(
  id = sample$id,
  y  = pred
)

write.csv(submission, "GAM4_Submission.csv", row.names=FALSE)

cat("Saved: GAM4_Submission.csv\n")
```
Score on Kaggle:
Private Score: 0.48633
Public Score: 0.49299